---
title: "Behind the Scenes"
author: "Jason Xu"
subtitle: "Slides updated from DS002 course"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Welcome...to the world of Macbeth

![](Macbeth){fig-align="center"}

Me in 6th grade:\
![](Jason%20Macbeth){fig-align="center"}

## The Data

"macbeth" from TidyTuesday!

```{r}
#| message: false
#| warning: false

library(dplyr)
library(tidyverse)

macbeth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-09-17/macbeth.csv')

head(macbeth)
```

## Negative Words

Which lines (or dialogue) contain negative words?\
I'll call it Negative Lines

```{r}
neg_lines <- macbeth |>
  mutate( blood = str_detect( dialogue, "(?i)blood|(?i)murder|(?i)death|(?i)kill" ) ) |>
  select( dialogue, blood)

head(neg_lines)
```

## Negative Lines Per Act

Total number of negative lines per act

```{r}
neg_lines <- macbeth |>
  mutate( blood = str_detect( dialogue, "(?i)blood|(?i)murder|(?i)death|(?i)kill" ) ) |>
  filter( blood ) |>
  group_by( act ) |>
  summarize( negative_lines_per_act = n() ) 

neg_lines
```

## String Length

String length counts the number of letters/characters/spaces in a string.

It is one way to quantify the length of each sentence of the dialogue.

```{r}
string_length = str_length( macbeth$dialogue )

head(string_length)                        
```

## Average String Length per Act

```{r}
average_character <- macbeth |>
  mutate( string_length = str_length( dialogue ) ) |>
  group_by( act ) |>
  summarize( average_character = mean( string_length ) ) 

average_character
```

## Is there a relationship?

Hypothesis: Characters in acts that contain more negative lines also tend to speak longer sentences (i.e longer string length).

```{r}
negative_words <- left_join( neg_lines, average_character )

head(negative_words)
```

## Is there a relationship?

-   <div>

    ![](NegWord%20Plot)

    </div>

-   Openning and closing (I and V) acts have the longest dialogues, but low negative lines

-   If any relationship: total negative lines per act seems to be negatively associated to average dialogue length per act -\> but comfounding variable: some acts may be longer than others

-   Act II has the most negative lines, but the shortest dialogues in average

## Limitations

-   "Negative" words are subjective! and only a very limited set of negative words were looked at here.\
-   Length of Acts were not captured: some acts may be much longer -\> leading to more negative words.

## What about "love"???

Does Macbeth ever verbally expressed "love" in the play??? and to whom?

```{r}
macbeth |>
  filter( character == "Macbeth" ) |>
  mutate( love_strings = str_extract( dialogue, "(?<=(?i)love ).+") ) |>
  filter( !is.na(love_strings) ) |>
  select( character:love_strings, -line_number )
```

Seems like it only occurred 4 times in the whole play and they seem to occur only in generic settings ("love and honour", "love and health to all", seems like words said when toasting) --- poor lady Macbeth :(

## References

Picture taken from: [https://medium.com/literary-analyses/lady-macbeth-30e549b7c211 Data taken from TidyTuesday](https://medium.com/literary-analyses/lady-macbeth-30e549b7c211%20Data%20taken%20from%20TidyTuesday)

GitHub Repository: <https://github.com/rfordatascience/tidytuesday/tree/main/data>\
Data originally sourced from: <https://shakespeare.mit.edu/>
