---
title: "Macbeth"
description: |
  Analysis of Shakespeare Dialogue Dataset from TidyTuesday
format: html
---

This dataset contains the entire script of the famous Shakesqeare play: Macbeth. As a side note, I was actually really into acting back in middle school, and I actually played a character (who's name I forgot - not a big character) in Macbeth during one of the school shows. I'm excited, let's go ahead and play with the data :)

```{r}
# Import Data
#| message: false
#| warning: false

library(dplyr)
library(tidyverse)

macbeth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-09-17/macbeth.csv')

# Total Number of Negative Lines in each act
neg_lines <- macbeth |>
  mutate( blood = str_detect( dialogue, "(?i)blood|(?i)murder[er]|(?i)death|(?i)kill" ) ) |>
  filter( blood ) |>
  group_by( act ) |>
  summarize( average_negative_lines = n() ) 

# Average Character per Line in each act
average_words <- macbeth |>
  mutate( string_length = str_length( dialogue ) ) |>
  group_by( act ) |>
  summarize( average_words = mean( string_length ) ) 

negative_words <- left_join( neg_lines, average_words )

ggplot( negative_words, aes(x = average_words, y = average_negative_lines, label = act)) +
  geom_point(color = "blue", size = 3) +  
  geom_text(vjust = -1, size = 4) +       
  labs(title = "Average Words vs. Negative Lines in each act of Macbeth",
       x = "Average Words per Act",
       y = "Number of Lines with Negative Words" ) +
  theme_minimal()
```

Here, I counted the number of lines in each act of Macbeth that contained one or more "negative words", which I defined to be "blood", "murder(er)", "death" and "kill". Then, it is paired up with the average number of words in each act (see figure above). I was interested in knowing if a particular act had more lines containing "negative words" than perhaps the characters spoke more on average (or less, there is some relationship is the point). However, just by eyeballing the above plot, it seems like there is no clear relationship between the two. (noticeably, however, Act III is a pretty extreme outlier, if we ignore it the plot actually shows a pretty down-wards trending relationship).

```{r}
#| message: false
#| warning: false
macbeth |>
  filter( character == "Macbeth" ) |>
  mutate( love_strings = str_extract( dialogue, "(?<=(?i)love ).+") ) |>
  filter( !is.na(love_strings) ) |>
  select( character:love_strings, -line_number )
```

After looking at the negative, I decided to turn to the opposite --- whether if Macbeth, the main character of the play, has explicity expressed "love" in the play, and to whom. So I used the lookaround function to spot instances of Macbeth saying the word "love" and the "love_strings" is whatever comes after "love". Sadly, these instances only occured 4 times in the whole play and they seem to occur only in more generic settings (like toasting) --- poor lady Macbeth :(

Data taken from TidyTuesday GitHub Repository: https://github.com/rfordatascience/tidytuesday/tree/main/data\
Data originally sourced from:\
https://shakespeare.mit.edu/
